Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/shell.tex
fixed shellsorts - numbers and documented errors with h = 5

Two versions of Gonnet's sort were used. The first was taken directly from the
code presented by Sedgewick, and only slightly modified to fit into the
framework used. The second was modified to reduce the complexity of the loop
bounds, and to remove a sentinel check in the inner loop. Code for shellsort and
improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
shellsort code}.

@@

Improved shellsort does improve on shellsort, having approximately 10\% fewer
instructions. It also has slightly fewer cache misses and branch mispredictions.
Overall, this leads to a reduction in cycle count of between 35\% and 70\%, as
seen in Figure \ref{shell cycles}.

@@

The results also show that in shellsort, a key moves only 1.03 places down the
array, on average, during each iteration. In improved shellsort, each key moves
over 1.07 places each iteration, while in the final insertion sort, each key
moves an average of 2.1 places.

@@

\section{Future Work}
Only Gonnet's way of producing increments is measured here. Using different
increments and different schemes for generating increments would produce
different results.

>>>>

Two versions of Gonnet's sort were used. The first was taken directly from the
code presented by Sedgewick, and only slightly modified to fit into the
framework used. The second was modified to reduce the complexity of the loop
bounds, and to remove a bounds check in the inner loop. Code for shellsort and
improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
shellsort code}.

A slight difference occurs in the two algorithms. Due to a slight difference
in behaviour when $h = 5$, the improved shellsort runs faster than shellsort
when this condition occurs. This is due to shellsort running an extra iteration,
with $h = 2$, which is skipped by improved shellsort. This occurs when the array
is sized $16,384$, $32,768$ and $4,194,304$.

@@

Improved shellsort does improve on shellsort, having approximately 10\% fewer
instructions. It also has slightly fewer cache misses and branch mispredictions.
Overall, this leads to a reduction in cycle count of between 35\% and 70\%, as
seen in Figure \ref{shell cycles}. While a small amount of this can be
attributed to the reduction in instruction count due to the removal of the
bounds check, a large amount cannot. We believe that the rest is due to slight
differences in the performance of the code generated by the compiler: register
allocation may be more efficient, for example, or better performing instruction
scheduling may take place.

@@

In general, the number of taken branches in both sorts would be almost exactly
the same, with only a minor difference due to the positioning of the sentinel.
However, during the sorts, the $h = 5$ condition was hit, meaning that there
were less branches and less taken branches in the improved shellsort, than in
the shellsort.

The results show that in improved shellsort, each key moves an average 0.915
places each iteration, while in the final insertion sort, each key moves an
average of 2.1 places. We can extrapolate a weighted average from this, to
determine that in shellsort, run under the same conditions, each key would move
an average of 0.9744.

@@

\section{Future Work}

Only Gonnet's way of producing increments is measured here. Using different
increments and different schemes for generating increments would produce
different results.

We note that improved shellsort performed in cases where $h = 5$ occurred.
Shellsort should also be investigated with this change made.
